---
title: "Kenya Economic Dashboard (2000-2023)"
author: "[Mocraig Kisali Sande]"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(plotly)
library(DT)
library(scales)

# Load data
kenya_data <- read_csv("../data/processed/kenya_economic_clean.csv")
```

## Executive Summary

This dashboard analyzes **23 years of Kenya's economic data (2000-2023)** from the World Bank, revealing a story of steady growth, trade challenges, and evolving macroeconomic stability.

**Key Findings:**
- GDP grew 8-fold from $14B to $113B
- Persistent trade deficit averaging -12% of GDP
- Inflation stabilized post-2012 after volatile period
- Rising debt-to-GDP ratio raises sustainability concerns

---

## 1. GDP Growth Trajectory

```{r gdp-growth}
p1 <- ggplot(kenya_data, aes(x = year, y = gdp_current_usd / 1e9)) +
  geom_line(color = "#006600", size = 1.5) +
  geom_point(color = "#006600", size = 2) +
  scale_y_continuous(labels = scales::dollar_format(suffix = "B")) +
  labs(
    title = "Kenya GDP (Current USD)",
    x = "Year",
    y = "GDP (Billions USD)",
    caption = "Source: World Bank WDI"
  ) +
  theme_minimal()

ggplotly(p1)
```

**Insight**: Kenya's economy expanded consistently, with only a brief contraction in 2020 (COVID-19).

---

## 2. GDP Growth Rate

```{r growth-rate}
p2 <- ggplot(kenya_data, aes(x = year, y = gdp_growth_annual)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_line(color = "#0066CC", size = 1.2) +
  geom_point(aes(color = gdp_growth_annual > 0), size = 2.5) +
  scale_color_manual(values = c("TRUE" = "#006600", "FALSE" = "#CC0000")) +
  labs(
    title = "Annual GDP Growth Rate",
    x = "Year",
    y = "Growth (%)",
    caption = "Source: World Bank WDI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p2)
```

**Average annual growth**: `r round(mean(kenya_data$gdp_growth_annual, na.rm = TRUE), 2)`%

---

## 3. Inflation Dynamics

```{r inflation}
p3 <- ggplot(kenya_data, aes(x = year)) +
  geom_line(aes(y = inflation_cpi), color = "gray50", size = 0.8) +
  geom_line(aes(y = inflation_ma3), color = "#FF6600", size = 1.5) +
  labs(
    title = "Inflation Rate (CPI)",
    subtitle = "Gray = Annual, Orange = 3-year moving average",
    x = "Year",
    y = "Inflation (%)"
  ) +
  theme_minimal()

ggplotly(p3)
```

**Insight**: High volatility (2008-2011) gave way to stability post-2015, reflecting improved monetary policy.

---

## 4. Trade Balance

```{r trade}
p4 <- ggplot(kenya_data, aes(x = year, y = trade_balance_pct_gdp)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_area(aes(fill = trade_balance_pct_gdp > 0), alpha = 0.3) +
  geom_line(size = 1) +
  scale_fill_manual(values = c("TRUE" = "#006600", "FALSE" = "#CC0000")) +
  labs(
    title = "Trade Balance (% of GDP)",
    x = "Year",
    y = "Balance (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p4)
```

**Persistent deficit**: Kenya has imported more than exported every year since 2000.

---

## 5. Debt Sustainability

```{r debt}
p5 <- ggplot(kenya_data, aes(x = year, y = debt_to_gdp_ratio)) +
  geom_line(size = 1.2, color = "#CC0000") +
  geom_point(size = 2, color = "#CC0000") +
  geom_hline(yintercept = 60, linetype = "dashed", color = "gray50") +
  labs(
    title = "External Debt-to-GDP Ratio",
    subtitle = "Dashed line = 60% sustainability threshold",
    x = "Year",
    y = "Debt-to-GDP (%)"
  ) +
  theme_minimal()

ggplotly(p5)
```

**Concern**: Debt-to-GDP crossed 60% threshold in 2018, requiring fiscal discipline.

---

## 6. Data Table

```{r data-table}
kenya_data %>%
  select(
    Year = year,
    `GDP (USD)` = gdp_current_usd,
    `GDP Growth (%)` = gdp_growth_annual,
    `Inflation (%)` = inflation_cpi,
    `Trade Balance (% GDP)` = trade_balance_pct_gdp,
    `Debt-to-GDP (%)` = debt_to_gdp_ratio
  ) %>%
  mutate(`GDP (USD)` = scales::dollar(`GDP (USD)`)) %>%
  datatable(
    options = list(pageLength = 10, scrollX = TRUE),
    caption = "Kenya Economic Indicators (2000-2023)"
  )
```

---

## Methodology

**Data Source**: World Bank World Development Indicators (WDI API)

**Indicators**: 10 core economic metrics including GDP, inflation, unemployment, trade, and debt.

**Processing**: Data cleaned for missing values, derived variables created (trade balance, debt ratios, moving averages).

**Tools**: R (tidyverse, plotly, DT), RMarkdown

**Repository**: [https://github.com/crayglockes/kenya-economic-dashboard]

---

## Contact

**Author**: [Mocraig Kisali Sande]  
**Email**: mocraigks@gmail.com  
**LinkedIn**: [linkedin.com/in/mocraigks]  
**GitHub**: [github.com/crayglockes]

*Built as part of a 14-day portfolio sprint demonstrating econometric analysis and data visualization skills.*